Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-17 treestats_paper.tex
Moved Duality in Practise and Dynamic Graph sections out of disco.

@@

@@

%%%%%%%%%%%%%%%%%%
\subsection*{Duality in practice?}

The duality between Site and Branch statistics depends crucially on two assumptions:
that each new mutation is distinguishable, as in the infinite-sites model,
and that mutation rates are constant along the genome.
Neither of these things are true, but are they good approximations?
Multiple mutations at the same site, causing back mutations,
are present in real datasets but are generally thought to be uncommon,
especially if hypervariable sites are excluded.
The effect of multiple mutations on Site statistics
can be bounded by a function of the proportion of sites at which these occur.
Large-scale variation in mutation rate (at, say, the megabase scale)
might plausibly have a stronger effect on the correspondence,
because a Branch and Site statistic are equal in expectation up to multiplication by the mutation rate.
For instance, if we knew that the inferred tree sequences of \autoref{fig:chr20} were correct,
then dividing the Site statistic by the Branch statistic
could give us an estimate of relative average mutation rate in windows along the genome.
Similarly, Branch statistics from well-calibrated tree sequences
would provide a way to assess variation in population genetic summaries along the genome
in a way independent of local mutation rate.

A final crucial assumption underlying this duality is that mutations can be treated
as independent of the trees themselves.
This is clearly not always true --- for instance,
a sweeping beneficial mutation strongly affects the local trees (as in \autoref{fig:sweep_duality}) ---
but seems likely to be approximately true.
\citet{ralph2019empirical} contains more discussion of variation in mutation rate,
but more work needs to be done, particularly on models with large proportions of sites under selection.

@@

%%%%%%%%%%%%%%%%%%
\subsection*{Dynamic graph algorithms}

The tree sequence describes trees that change as one moves along the genome,
and so is a special case of a \emph{dynamic graph},
also called a \emph{graph timeline} \citep{lacki2013reachability}.
Much of the work on dynamic graphs is focused on connectivity,
e.g., maintaining a minimum spanning tree \citep{eppstein1994offline, eppstein1997sparsification, holm2001polylogarithmic},
but development of parallel algorithms for more general operations on large dynamic graphs
is an active area of research \citep[e.g.,][]{srinivasan2018sharedmemory}.

>>>>

% This paragraph is essentially saying the same thing as the previous one,
% so am cutting it. Perhaps we could merge some of the content into the
% the paragtraph above?

% The duality between Site and Branch statistics depends crucially on two assumptions:
% that each new mutation is distinguishable, as in the infinite-sites model,
% and that mutation rates are constant along the genome.
% Neither of these things are true, but are they good approximations?
% Multiple mutations at the same site, causing back mutations,
% are present in real datasets but are generally thought to be uncommon,
% especially if hypervariable sites are excluded.
% The effect of multiple mutations on Site statistics
% can be bounded by a function of the proportion of sites at which these occur.
% Large-scale variation in mutation rate (at, say, the megabase scale)
% might plausibly have a stronger effect on the correspondence,
% because a Branch and Site statistic are equal in expectation up to multiplication by the mutation rate.
% For instance, if we knew that the inferred tree sequences of \autoref{fig:chr20} were correct,
% then dividing the Site statistic by the Branch statistic
% could give us an estimate of relative average mutation rate in windows along the genome.
% Similarly, Branch statistics from well-calibrated tree sequences
% would provide a way to assess variation in population genetic summaries along the genome
% in a way independent of local mutation rate.

A final crucial assumption underlying the duality between Site and Branch
statistics is that mutations can be treated as independent of the trees themselves.
This is clearly not always true --- for instance,
a sweeping beneficial mutation strongly affects the local trees (as in \autoref{fig:sweep_duality}) ---
but seems likely to be approximately true.
\citet{ralph2019empirical} contains more discussion of variation in mutation rate,
but more work needs to be done, particularly on models with large proportions of sites under selection.

@@

The tree sequence describes trees that change as one moves along the genome,
and so is a special case of a \emph{dynamic graph},
also called a \emph{graph timeline} \citep{lacki2013reachability}.
Much of the work on dynamic graphs is focused on connectivity,
e.g., maintaining a minimum spanning tree \citep{eppstein1994offline, eppstein1997sparsification, holm2001polylogarithmic},
but development of parallel algorithms for more general operations on large dynamic graphs
is an active area of research \citep[e.g.,][]{srinivasan2018sharedmemory}.
An interesting direction for future work is to develop
parallel versions of tree sequence algorithms such as Algorithm B.

@@

%%%%%%%%%%%%%%%%%%
% \subsection*{Duality in practice?}
% Moved this to the 1000G section where we're already discussing it.

@@

%%%%%%%%%%%%%%%%%%
% Moved this content into the "Algorithms" section
% \subsection*{Dynamic graph algorithms}
